---
title: "WEB DuBois Challenge"
author: "Marieke Jones"
date: "2/16/2021"
output: 
  html_document:
    keep_md: true
---

# Load the data

```{r message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
```

Try to re-create WEB Du Bois's [original income figure](https://github.com/ajstarks/dubois-data-portraits/blob/master/challenge/challenge05/original-plate-31.jpg)

Read in the income data
```{r}
income <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-16/income.csv')
income
```

Looks like the original plots the types of expenses so we should `pivot_longer()`

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income")
```

Yes, that is better

Now try a basic plot

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  ggplot(aes(Percent_of_Income, Class)) + 
  geom_col(aes(fill = Expense))
```

Already close to the original, but the y-axis is upside down, so reverse that

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  ggplot(aes(Percent_of_Income, Class)) + 
  geom_col(aes(fill = Expense)) +
  scale_y_discrete(limits = rev)
```

Change the order of the expenses and the class to match those in the original

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense))

# now plot
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense)) %>%
  ggplot(aes(Percent_of_Income, Class)) + 
  geom_col(aes(fill = Expense)) +
  scale_y_discrete(limits = rev)
```

Change the colors to match the original

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense)) %>%
  ggplot(aes(Percent_of_Income, Class)) + 
  geom_col(aes(fill = Expense)) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = c("azure2", "steelblue3", "lightpink", "mediumpurple4", "black"))
```

Change the spacing between the bars and remove axes labels and change theme

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense)) %>%
  ggplot(aes(Percent_of_Income, Class)) + 
  geom_col(aes(fill = Expense), width = .4) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = c("azure2", "steelblue3", "lightpink", "mediumpurple4", "black")) +
  theme_classic() +
  labs(x = "", y = "")
```

Add annotations in each segment. To do so, replace NA in Percent of Income variable with 0 first

```{r}
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense)) %>%
  mutate(Percent_of_Income = replace_na(Percent_of_Income, 0))

# now plot
income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense)) %>%
  mutate(Percent_of_Income = replace_na(Percent_of_Income, 0)) %>%
  ggplot(aes(Percent_of_Income, Class)) + 
  geom_col(aes(fill = Expense), width = .4) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = c("azure2", "steelblue3", "lightpink", "mediumpurple4", "black")) +
  theme_classic() +
  labs(x = "", y = "") +
  geom_text(aes(label = paste0(Percent_of_Income, " %")), size = 3, position = position_stack(vjust = .5), color = rep(c("white", rep("black", 4)), 7))
```

Add the lines connecting each portion of the bar and finalize details (background color, title, etc.)

```{r}
plotdata <- income %>%
  pivot_longer(Rent:Other, names_to = "Expense", values_to = "Percent_of_Income") %>%
  mutate(Expense = factor(Expense, levels = c("Other", "Tax", "Clothes", "Food", "Rent"))) %>%
  mutate(Class = factor(Class, levels = c("$100-200", "$200-300", "$300-400", "$400-500", "$500-750", "$750-1000", "Over $1000"))) %>%
  arrange(Class, desc(Expense)) %>%
  mutate(Percent_of_Income = replace_na(Percent_of_Income, 0)) 
plotdata
  
# create cumsum column and make cumsum NA for Other
plotdata <- plotdata %>%
  group_by(Class) %>%
  mutate(Cumulative_perc = cumsum(Percent_of_Income)) %>% ungroup() %>%
  mutate(Cumulative_perc = ifelse(Expense == "Other", NA, Cumulative_perc))

# take a look
plotdata

# now plot
plotdata %>%
  ggplot() + 
  geom_col(aes(Percent_of_Income, Class, fill = Expense), width = .4) +
  scale_y_discrete(limits = rev) +
  geom_path(aes(Cumulative_perc, Class, group = Expense), lwd = .2) +
  scale_fill_manual(values = c("azure2", "steelblue3", "lightpink", "mediumpurple4", "black")) +
  theme_classic() +
  labs(x = "", y = "") +
  geom_text(aes(Percent_of_Income, Class, label = paste0(Percent_of_Income, " %")), size = 3, position = position_stack(vjust = .5), color = rep(c("white", rep("black", 4)), 7)) + 
  theme(
  plot.background=element_rect(fill = "antiquewhite"),
  panel.background = element_rect(fill = 'antiquewhite'),
  legend.background = element_rect(fill = "antiquewhite", color = NA),
  axis.line = element_blank()) +
  labs(title = "INCOME AND EXPENDITURE OF 150 NEGRO FAMILIES IN ATLANTA, GA., U.S.A.", caption = "reproducing WEB DuBois's original data graphic using ggplot2")

# save the final plot
ggsave(filename = "dubois_challenge.png", height = 8, width = 12)
```
